---
title: ""
subtitle: "JSC370 Midterm Report"
author: "Shih-Ting (Cindy) Huang"
date: "02/27/2022"
output: 
  html_document:
    theme: lumen
    highlight: tango
---

## Introduction 
Research question: How has COVID-19 affected restaurant performance?
* background on dataset and formulated question

Yelp is one of the biggest platforms that shares information of businesses and enable users to search for, react to, and/or share their opinions of such businesses. After the emergence of COVID-19 in early 2020, undoubtedly many businesses were forced to close temporarily or permanently. 


## Methods
The data were retrieved directly from [Yelp's Open Dataset](https://www.yelp.com/dataset) in a `.tar` compressed file. Extracting the respective json files and converting each to csv using Python, we select two relevant datasets - `business` and `review`, which provides details on a business and a review respectively. Data comes from 6,990,280 reviews and 150,346 businesses from 11 metropolitan areas. 

All figures were created with `ggplot2`.

* how and where the data were acquired
* how you cleaned and wrangled the data
* what tools you used for data exploration

## Preliminary Results
* summary statistics

## Conclusion
* what i found in terms of the question
```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(leaflet)
```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
business <- data.table::fread("yelp_business.csv")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
head(business)
```





















